% JCSSE Bangkok 2026 - Full Paper Template (Regular/Special Session)
% The 23rd International Joint Conference on Computer Science and Software Engineering
% 24-27 June 2026, Bangkok, Thailand
% https://jcsse2026.org/
%
% IEEE Conference Proceedings format - A4 paper, Times New Roman 10pt
% Full Paper: up to 6 pages
% BLIND REVIEW: Do NOT include author names/affiliations until camera-ready

\documentclass[10pt,twocolumn,a4paper]{article}

% ===== PACKAGES =====
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{times}                      % Times New Roman font
\usepackage{mathptmx}                   % Times for math
\usepackage{graphicx}                   % For figures
\usepackage{amsmath,amssymb}            % Math support
\usepackage{caption}                    % Caption styling
\usepackage{array}                      % Table improvements
\usepackage{booktabs}                   % Professional tables
\usepackage{enumitem}                   % List customization
\usepackage{hyperref}                   % Hyperlinks (disabled for IEEE style)
\usepackage{xcolor}                     % Color support
\usepackage{balance}                    % Balance last page columns
\usepackage{titlesec}                   % Section formatting
\usepackage{fancyhdr}                   % Header/footer control
\usepackage{flushend}                   % Balance final page columns
\usepackage{algorithm}                  % Algorithm environment
\usepackage{algorithmic}                % Algorithm formatting
\usepackage{cite}                       % Citation handling
\usepackage{tikz}                       % Diagrams
\usetikzlibrary{arrows.meta,positioning,calc,shapes.geometric}

% ===== PAGE LAYOUT (A4 Paper - IEEE Standard) =====
% A4: 210mm x 297mm
% Top margin: 19mm, Bottom: 43mm, Left/Right: 13mm
% Column width: 88mm, Column gap: 6mm
\usepackage[
    a4paper,
    top=19mm,
    bottom=43mm,
    left=13mm,
    right=13mm,
    columnsep=6mm
]{geometry}

% ===== NO PAGE NUMBERS =====
\pagestyle{empty}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}

% ===== HYPERREF SETUP (disable visible links per IEEE style) =====
\hypersetup{
    colorlinks=false,
    pdfborder={0 0 0},
    hidelinks
}

% ===== SECTION FORMATTING =====
% First-order headings: 10pt small caps, centered, 12pt before, 6pt after
\titleformat{\section}
    {\normalfont\normalsize\scshape\centering}
    {}
    {0pt}
    {}
\titlespacing*{\section}{0pt}{12pt}{6pt}

% Second-order headings: 10pt italic, flush left, Roman numerals, 9pt before, 3pt after
\titleformat{\subsection}
    {\normalfont\normalsize\itshape}
    {\Roman{subsection}.}
    {0.5em}
    {}
\titlespacing*{\subsection}{0pt}{9pt}{3pt}

% Third-order headings: 10pt italic, 6pt before, 3pt after
\titleformat{\subsubsection}
    {\normalfont\normalsize\itshape}
    {\alph{subsubsection})}
    {0.5em}
    {}
\titlespacing*{\subsubsection}{0pt}{6pt}{3pt}

% Reset numbering
\renewcommand{\thesubsection}{\Roman{subsection}}
\renewcommand{\thesubsubsection}{\alph{subsubsection}}

% ===== CAPTION FORMATTING =====
% 10pt Times, small caps, left-justified
\captionsetup{
    font=small,
    labelfont={sc},
    justification=justified,
    singlelinecheck=false,
    skip=6pt
}
\captionsetup[figure]{position=below}
\captionsetup[table]{position=above}

% ===== PARAGRAPH FORMATTING =====
\setlength{\parindent}{1pc}             % 1 pica indent (~0.17 inch)
\setlength{\parskip}{0pt}               % No space between paragraphs

% ===== LIST FORMATTING =====
\setlist[itemize]{
    leftmargin=0.5in,
    topsep=0pt,
    partopsep=0pt,
    parsep=0pt,
    itemsep=0pt
}
\setlist[enumerate]{
    leftmargin=0.5in,
    topsep=0pt,
    partopsep=0pt,
    parsep=0pt,
    itemsep=0pt
}

% ===== CUSTOM COMMANDS =====
% Abstract environment (10pt bold italic)
\renewenvironment{abstract}{%
    \noindent\textbf{\textit{Abstract}}\textemdash\textbf{\textit{}}%
}{%
    \par\vspace{6pt}
}

% Index Terms
\newcommand{\indexterms}[1]{%
    \noindent\textbf{\textit{Index Terms}}\textemdash\textit{#1}%
    \par\vspace{12pt}
}

% Source for figures/tables
\newcommand{\source}[1]{%
    \vspace{1pt}
    {\footnotesize Source: #1}
}

% ===== DOCUMENT =====
\begin{document}

% ===== TITLE =====
\twocolumn[
\begin{center}
    {\fontsize{24pt}{28pt}\selectfont Explainable Adversarial Drift Detection\\[4pt] for MLOps Feature Monitoring\par}
    \vspace{18pt}

    % ===== BLIND REVIEW VERSION =====
    % IMPORTANT: For initial submission, DO NOT include author information.
    % Uncomment the author block below ONLY for camera-ready submission:
    % -----------------------------------------------------------------------
    % \begin{tabular}{ccc}
    %     \begin{tabular}{c}
    %         \textbf{Nusrat Begum} \\
    %         \textit{Faculty of ICT, Mahidol University} \\
    %         nusrat.beg@student.mahidol.edu
    %     \end{tabular}
    %     &
    %     \begin{tabular}{c}
    %         \textbf{Thanapon Noraset} \\
    %         \textit{Faculty of ICT, Mahidol University} \\
    %         thanapon.nor@mahidol.edu
    %     \end{tabular}
    %     &
    %     \begin{tabular}{c}
    %         \textbf{Suppawong Tuarob} \\
    %         \textit{Faculty of ICT, Mahidol University} \\
    %         suppawong.tua@mahidol.edu
    %     \end{tabular}
    % \end{tabular}
    % \vspace{18pt}
    % -----------------------------------------------------------------------

\end{center}
]

% ===== ABSTRACT (~150 words) =====
\noindent\textbf{\textit{Abstract}}\textemdash\textit{Machine learning models in production suffer from distribution drift, where evolving input features degrade model performance. Existing unsupervised detectors signal that drift occurred but not which features drifted or what corrective action to take. This paper proposes Explainable Adversarial Drift Detection (EADD), extending discriminative classification with four novel contributions: (1) prediction entropy monitoring for early warning of emerging distributional instability, (2) an Adaptive Sequential Permutation Test (ASPT) that reduces computational overhead by up to 70\% while maintaining Type~I error control, (3) SHAP interaction-aware drift localisation for correlated feature diagnosis, and (4) a Drift Severity Index (DSI) for continuous severity quantification and automated remediation prescriptions. Evaluated against seven state-of-the-art detectors across synthetic and 11 real-world benchmarks, EADD uniquely achieved complete drift-type coverage (4/4) with near-zero false alarms while providing perfect feature-level attribution (P@k\,{=}\,R@k\,{=}\,1.0).}

\vspace{6pt}

% ===== INDEX TERMS =====
\noindent\textbf{\textit{Index Terms}}\textemdash\textit{concept drift, discriminative classification, feature attribution, distribution shift, MLOps, explainable AI.}

\vspace{12pt}

% ===== INTRODUCTION =====
\section{Introduction}

Machine learning models deployed in production environments encounter evolving data distributions. These changes---collectively termed distribution drift---can manifest as changes in input features (covariate drift) or in the input-output relationship (concept drift), causing performance degradation if left unmanaged~[1],~[2]. Detecting such changes early is essential for maintaining reliable inference~[3],~[4].

Conventional monitoring strategies react to drift only after observable performance degradation, which is inadequate in production contexts where ground-truth labels may be delayed or unavailable~[1],~[5]. The drift detection landscape has matured considerably: statistical methods employ hypothesis tests such as the Kolmogorov--Smirnov test~[6],~[7]; supervised detectors like DDM monitor error rates but require labels~[8]; and discriminative methods like D3 train classifiers to distinguish old from new data, achieving state-of-the-art accuracy in recent benchmarks~[9],~[10]. However, all existing methods share a common limitation: they report only \textit{whether} drift occurred, not \textit{which} features drifted or \textit{what} corrective action is appropriate.

Recent advances have begun addressing individual aspects of this gap, but each solves only one dimension, leaving the integration gap unresolved. E-C2ST~[27] specifically targets the \textit{statistical rigour gap} in D3: it replaces D3's fixed-budget permutation test with anytime-valid e-values, enabling sequential testing with principled Type~I error control. However, E-C2ST retains D3's detect-and-discard pattern---it still provides no feature-level attribution. Practitioners using E-C2ST learn \textit{whether} drift is statistically confirmed, but not \textit{which} features are responsible or how severe the shift is. TRIPODD~[28]\footnote{Feature-inTeraction awaRe InterPretable mOdel Drift Detection.} specifically targets the \textit{feature identification gap}: it introduces interaction-aware hypothesis testing that can identify which feature pairs drive drift. However, TRIPODD requires pre-specifying interaction hypotheses and relies on conditional independence tests with quadratic complexity in the number of features, making it impractical for real-time monitoring of high-dimensional streams; it also provides no automated remediation guidance. HDPD~[29] targets the \textit{attribution accuracy gap} by decomposing performance discrepancy into per-feature contributions with nonparametric guarantees. However, HDPD requires ground-truth labels (it is supervised) and operates post-hoc on collected batches, making it inapplicable to unsupervised, real-time streaming contexts.

These partial solutions leave a critical operational gap. In production, distribution shifts are not isolated events: a feature may drift subtly for weeks before a statistical detector triggers, at which point the practitioner still does not know \textit{which} of potentially hundreds of features caused it, how severe the shift is, or whether to retrain the model, source more data, or flag a specific data pipeline. Manual diagnosis of individual features is impractical at scale: the time between a drift alarm and remediation (MTTD to MTTR) directly translates to degraded predictions and, in regulated domains such as healthcare or credit scoring, potential compliance violations under frameworks such as the EU AI Act~[3],~[4]. Production ML systems therefore require a single integrated pipeline that can: (1) detect drift with statistical rigour and computational efficiency, (2) identify \textit{which} specific features and feature interactions caused the shift, (3) quantify \textit{how severe} the drift is relative to the baseline, and (4) automatically recommend a \textit{proportional corrective action}---all without ground-truth labels. Without this integration, practitioners must combine multiple disconnected tools, manually correlate their outputs, and interpret conflicting signals, significantly delaying remediation and increasing the risk of silent model degradation~[1],~[4].

This paper proposes \textbf{Explainable Adversarial Drift Detection (EADD)}, addressing this gap with four contributions:
\begin{enumerate}
    \item \textbf{Prediction entropy monitoring} of the discriminative classifier, providing early-warning signals for emerging distributional instabilities before statistical significance is reached.
    \item An \textbf{Adaptive Sequential Permutation Test (ASPT)}~[30] that reduces computational overhead by up to 70\% while maintaining Type~I error control.
    \item \textbf{SHAP interaction-aware drift localisation}~[31] that identifies correlated feature drift invisible to marginal attribution.
    \item A \textbf{Drift Severity Index (DSI)} for continuous severity quantification and automated, proportional remediation prescriptions.
\end{enumerate}

The remainder of this paper is organized as follows: Section~II reviews related work. Section~III describes the EADD methodology. Section~IV presents experimental evaluation. Section~V concludes.

% ===== RELATED WORK =====
\section{Related Work}

\subsection{Drift Detection Methods}

Drift detection methods can be classified along two axes: label availability and detection mechanism~[1],~[2].

\textit{Supervised methods} monitor model error rates and require timely ground-truth labels. DDM~[8] tracks the online error rate and its standard deviation, triggering drift alerts when these statistics exceed binomial distribution bounds. EDDM~[11] extends DDM by monitoring the distance between consecutive errors, improving sensitivity to gradual drift. Both methods react to drift only \textit{after} performance has already degraded, limiting their utility in production settings where labels may be delayed by days or weeks~[1],~[5].

\textit{Unsupervised methods} operate without labels by directly monitoring data distributions. Statistical approaches include ADWIN~[12], which dynamically adjusts sliding window sizes using statistical change tests, and KSWIN~[7], which applies the Kolmogorov--Smirnov two-sample test over sliding windows. While statistically principled, these methods test each feature independently (univariate), leading to multiple testing problems in high-dimensional settings~[6]. Discriminative approaches instead train auxiliary classifiers to distinguish between reference and current distributions, framing drift detection as binary classification~[9],~[13],~[14],~[19]. This naturally handles multivariate drift in a single unified test. The most prominent discriminative detector is D3~[9], which trains a logistic regression classifier on reference vs.\ current windows and triggers drift when the out-of-fold AUC exceeds a fixed threshold of 0.7. D3 was identified as the most robust unsupervised detector in the comprehensive benchmark by Lukats et al.~[10]. However, D3 has three critical weaknesses that limit its operational utility: (1)~it uses a fixed AUC threshold calibrated on specific benchmark conditions, making it vulnerable to false alarms on correlated or autocorrelated data; (2)~it applies no formal significance test, so a borderline AUC of 0.71 is treated identically to a clear AUC of 0.95; and (3)~it discards the trained classifier after computing AUC, providing no information about \textit{which} features drove the detected drift. Other benchmarked detectors exhibit characteristic limitations: CSDDM~[15] (PCA + K-Means clustering) and IBDD~[16] (image-based comparison) achieve high coverage but suffer elevated false alarm rates; OCDD~[17] (one-class SVM) and SPLL~[18] (semi-parametric log-likelihood) are effective for specific drift types but miss others; BNDM (Bayesian non-parametric) struggles with small window sizes; and UDetect (Shewhart control charts) failed on 45\% of benchmark scenarios~[10].

\subsection{Identifying Drift-Causing Features}

The \textit{classifier two-sample test} (C2ST), formalised by Lopez-Paz and Oquab~[19], frames distribution comparison as binary classification: data from a reference period is labelled 0 and data from the current period is labelled 1; if a classifier can reliably separate them (AUC $\gg 0.5$), the distributions differ significantly. This approach has been adopted at Uber for detecting train-deployment mismatch~[13] and for credit scoring shift management~[14]. D3~[9] applies this principle in streaming settings using logistic regression as the discriminator with a fixed AUC threshold---effectively a C2ST without formal significance testing.

Despite strong detection accuracy, existing C2ST-based detectors---including D3---discard the trained classifier after computing AUC, losing information about \textit{which} features differentiate the distributions. This ``detect-and-discard'' pattern wastes the diagnostic value embedded in the classifier's learned decision boundary~[19]. Industrial evidence validates the need for feature-level attribution: Google's Vertex AI uses SHAP-based feature attribution monitoring that detected degradation missed by aggregate input-distribution monitoring~[20], and Amazon SageMaker Clarify similarly integrates SHAP-based attribution for production drift detection~[21]. Recent academic work has advanced individual aspects: TRIPODD~[28] enables feature-interaction aware hypothesis testing, HDPD~[29] provides hierarchical performance decomposition with per-feature attributions, and Hinder et al.~[32] formalised rigorous mathematical definitions of feature-wise drift. However, no existing work systematically evaluates the accuracy of SHAP-based feature attribution for identifying which features cause drift within discriminative detection frameworks---nor integrates such attribution with severity quantification and automated prescriptions in an unsupervised streaming pipeline.

% ===== METHODOLOGY =====
\section{Proposed Methodology}

As identified in Section~II, existing C2ST-based detectors lack early-warning signals, statistical significance testing, feature-level attribution, and severity quantification. EADD is a streaming pipeline for feature drift ($P(X)$ shift) detection that extends the C2ST framework~[19] with four novel components to address each of these gaps (Fig.~\ref{fig:pipeline}).

\begin{figure}[htbp]
    \centering
    \begin{tikzpicture}[
        node distance=0.5cm,
        box/.style={rectangle, draw, rounded corners=2pt, minimum width=2.1cm, minimum height=0.55cm, align=center, font=\scriptsize},
        decision/.style={diamond, draw, aspect=2, minimum width=0.8cm, align=center, font=\scriptsize},
        arrow/.style={-{Stealth[scale=0.6]}, thick},
        novel/.style={box, fill=blue!12, draw=blue!50},
        standard/.style={box, fill=gray!8},
        lbl/.style={font=\tiny}
    ]
    \node[standard] (stream) {Data Stream\\$\{x_t\}$};
    \node[standard, right=0.4cm of stream] (window) {Step 1:\\Windowing};
    \node[standard, right=0.4cm of window] (adv) {Step 2:\\C2ST\\(LightGBM)};
    \node[novel, below=0.4cm of adv] (pui) {\textbf{$H_{\text{pred}}$}\\(Novel)};
    \node[decision, right=0.4cm of adv] (dec1) {\tiny AUC\\$>\tau$?};
    \node[novel, right=0.4cm of dec1] (perm) {Step 3:\\\textbf{ASPT}\\(Novel)};
    \node[decision, below=0.6cm of perm] (dec2) {\tiny $p<$\\$\alpha$?};
    \node[novel, left=0.4cm of dec2] (shap) {Step 4:\\\textbf{SHAP+DSI}\\(Novel)};
    \node[novel, left=0.4cm of shap] (output) {Drift Report\\+ Prescription};
    \node[standard, below=0.8cm of dec1] (nodrift) {No Drift};
    \draw[arrow] (stream) -- (window);
    \draw[arrow] (window) -- (adv);
    \draw[arrow] (adv) -- (pui);
    \draw[arrow] (adv) -- (dec1);
    \draw[arrow] (dec1) -- node[above, lbl] {Yes} (perm);
    \draw[arrow] (dec1) -- node[right, lbl] {No} (nodrift);
    \draw[arrow] (perm) -- (dec2);
    \draw[arrow] (dec2) -- node[above, lbl] {Yes} (shap);
    \draw[arrow] (dec2.south) -- ++(0,-0.3) -| node[near start, right, lbl] {No} (nodrift.south);
    \draw[arrow] (shap) -- (output);
    \end{tikzpicture}
    \caption{EADD pipeline overview. Shaded boxes indicate the four novel contributions beyond D3~[9]: Prediction Entropy ($H_{\text{pred}}$, Step~2), Adaptive Sequential Permutation Test (ASPT, Step~3), and SHAP interaction-aware attribution with Drift Severity Index (DSI, Step~4).}
    \label{fig:pipeline}
\end{figure}

\subsection{Step 1: Adaptive Reference Windowing}

EADD maintains two data windows. The \textit{reference window} $W_{ref}$ ($N_{ref}$ samples) represents the baseline distribution and is maintained via \textit{reservoir sampling}~[22]: each new sample $x_t$ replaces a randomly chosen element in $W_{ref}$ with probability $N_{ref}/t$, ensuring that all past samples have equal probability of representation regardless of stream length. This is critical because a simple sliding reference window would only capture the most recent history, not the global pre-drift distribution. The \textit{current window} $W_{cur}$ ($N_{cur}$ samples) is a standard sliding window containing the most recent observations. Monitoring is triggered every $M$ samples to balance detection latency with computational cost. Upon confirmed drift, $W_{ref}$ is reset to the current data to establish a new baseline.

\subsection{Step 2: Discriminative Classification and Prediction Entropy}

Following the C2ST framework~[19], EADD constructs a binary classification task: samples from $W_{ref}$ are labelled 0 (reference) and samples from $W_{cur}$ are labelled 1 (current). A LightGBM gradient boosting classifier~[23] is trained on this combined dataset via stratified 5-fold cross-validation, and the out-of-fold AUC-ROC is computed. AUC~$\approx$~0.5 indicates the classifier cannot distinguish the two windows (no drift); AUC~$\gg$~0.5 indicates the distributions are separable (drift suspected). We use LightGBM rather than D3's logistic regression~[9] because gradient boosting captures non-linear feature interactions and has been shown to be a more powerful discriminator in two-sample testing~[19].

\textbf{Prediction Entropy.} Beyond the aggregate AUC score, EADD extracts additional diagnostic information from the classifier's per-sample predictions. We compute the normalised binary entropy of the discriminative classifier's predicted probabilities:
\begin{equation}
H_{\text{pred}} = \frac{-1}{n}\sum_{i=1}^{n}\left[\hat{p}_i\log\hat{p}_i + (1{-}\hat{p}_i)\log(1{-}\hat{p}_i)\right]
\label{eq:pui}
\end{equation}
where $\hat{p}_i = f_\theta(x_i)$ is the predicted probability that sample $i$ belongs to $W_{cur}$. $H_{\text{pred}} \approx \ln 2$ indicates maximum uncertainty (no drift); $H_{\text{pred}} \rightarrow 0$ indicates high classifier confidence (drift likely). A declining $H_{\text{pred}}$ trend across consecutive windows serves as an early-warning signal for emerging drift before statistical significance is reached.

\subsection{Step 3: Adaptive Sequential Permutation Test (ASPT)}

If AUC exceeds a threshold $\tau$, EADD applies an \textbf{adaptive sequential permutation test} inspired by Gandy~[30]. Instead of a fixed budget of $B$ permutations, ASPT monitors evidence accumulation and terminates early:
\begin{enumerate}
    \item For each permutation $i = 1, \ldots, B_{\max}$: shuffle labels, retrain, compute $\text{AUC}^{(i)}_{\text{perm}}$. Update count $c_i = \#\{j \leq i : \text{AUC}^{(j)}_{\text{perm}} \geq \text{AUC}_{\text{actual}}\}$.
    \item \textbf{Early reject $H_0$:} If $c_i = 0$ and $i \geq B_{\min}$ and $(1{-}\alpha)^i < 0.05$, terminate and confirm drift ($p = 0.001$).
    \item \textbf{Early accept $H_0$:} If $c_i/i > 2\alpha$ and $i \geq B_{\min}$, terminate and declare no drift ($p = 1.0$).
    \item \textbf{Terminal:} If $B_{\max}$ is reached, $p = (c_{B_{\max}}{+}1)/(B_{\max}{+}1)$.
\end{enumerate}
Drift is confirmed if $p < \alpha$. Gandy~[30] proved that early stopping maintains Type~I error control when termination rules satisfy the bounded resampling risk condition, enabling substantial computational savings without sacrificing statistical validity.

\subsection{Step 4: Feature Attribution and Drift Severity Index}

Upon drift confirmation, EADD exploits the trained discriminative classifier---which D3~[9] would discard---to diagnose \textit{which} features drive the detected drift using SHAP (SHapley Additive exPlanations)~[24].

\textbf{Feature Attribution.} EADD applies TreeSHAP~[24], a polynomial-time algorithm for computing exact Shapley values for tree-based models, to the trained LightGBM discriminator. For each sample, TreeSHAP produces a vector of SHAP values $\phi_j$ for each feature $j \in \{1, \ldots, d\}$ (where $d$ is the number of input features), indicating how much that feature contributed to the classifier's ability to distinguish reference from current data. Marginal SHAP importances are computed as $\bar{\phi}_j = \frac{1}{n}\sum_{i=1}^{n}|\phi_j^{(i)}|$ and normalised to obtain a contribution vector $\mathbf{s} = (s_1, \ldots, s_d)$ where $s_j = \bar{\phi}_j / \sum_k \bar{\phi}_k$. This vector serves two purposes: (1)~it directly identifies which individual features drive the detected drift, and (2)~it is used to compute drift severity via the DSI (below). Additionally, EADD computes \textbf{SHAP interaction values}~[31] via the interaction matrix $\Phi \in \mathbb{R}^{d \times d}$, where $\Phi_{ij}$ captures the joint effect of features $i$ and $j$ beyond their individual contributions. Significant interactions (exceeding $\mu + 2\sigma$ of the interaction distribution) are reported as supplementary diagnostics, addressing \textit{correlated feature drift}---where the joint distribution $P(X_i, X_j)$ shifts while the marginals $P(X_i)$ and $P(X_j)$ remain stable---a pattern that marginal attribution alone misses~[32]. The DSI below uses the marginal vector $\mathbf{s}$ for severity quantification, while interaction diagnostics guide practitioners toward correlated root causes.

\textbf{Drift Severity Index (DSI).} Using the marginal contribution vector $\mathbf{s}$, EADD computes a composite severity metric combining detection confidence with attribution concentration:
\begin{equation}
\text{DSI} = \text{AUC} \times \left(1 - H_{\text{norm}}(\mathbf{s})\right)
\label{eq:dsi}
\end{equation}
where $\mathbf{s} = (s_1, \ldots, s_d)$ is the normalised SHAP contribution vector (i.e., $s_j = \bar{\phi}_j / \sum_k \bar{\phi}_k$), $d$ is the number of input features, and $H_{\text{norm}}(\mathbf{s}) = -\sum_{j=1}^{d} s_j \log_2 s_j \,/\, \log_2 d$ is the normalised Shannon entropy~[34]. The $\log_2 d$ denominator ensures $H_{\text{norm}} \in [0, 1]$ regardless of dimensionality. Intuitively, when drift is concentrated in a single feature, $H_{\text{norm}} \approx 0$ and $\text{DSI} \approx \text{AUC}$ (high severity); when drift is uniformly spread across all features, $H_{\text{norm}} \approx 1$ and $\text{DSI} \approx 0$ (diffuse shift). DSI-based prescriptions replace heuristic thresholds:
\begin{itemize}
    \item \textbf{Critical (DSI $> 0.6$):} Concentrated drift $\rightarrow$ audit specific data pipeline.
    \item \textbf{Moderate ($0.3$--$0.6$):} Subset drift $\rightarrow$ investigate shared data source.
    \item \textbf{Low ($< 0.3$):} Diffuse shift $\rightarrow$ schedule full model retraining.
\end{itemize}

\subsection{Algorithm Summary}

\begin{algorithm}[htbp]
\caption{EADD Detection Cycle}
\label{alg:eadd}
\begin{algorithmic}[1]
\REQUIRE Stream $\{x_t\}$, $W_{ref}$, $W_{cur}$, threshold $\tau$, significance $\alpha$
\ENSURE Drift decision, $H_{\text{pred}}$, DSI, feature attribution, prescription
\STATE Update $W_{cur}$; update $W_{ref}$ via reservoir sampling
\STATE Train LightGBM on $W_{ref} \cup W_{cur}$ with labels $\{0,1\}$
\STATE Compute AUC via stratified 5-fold CV
\STATE Compute $H_{\text{pred}}$ via Eq.~(\ref{eq:pui})
\IF{AUC $> \tau$}
    \STATE $p \gets$ ASPT($W_{ref}$, $W_{cur}$, AUC) \COMMENT{Adaptive early stopping}
    \IF{$p < \alpha$}
        \STATE Compute SHAP values + interaction matrix via TreeSHAP
        \STATE Compute DSI via Eq.~(\ref{eq:dsi})
        \STATE Generate DSI-based prescription
        \RETURN \textsc{Drift}, $H_{\text{pred}}$, DSI, SHAP ranking, prescription
    \ENDIF
\ENDIF
\RETURN \textsc{No-Drift}, $H_{\text{pred}}$
\end{algorithmic}
\end{algorithm}

% ===== EXPERIMENTAL EVALUATION =====
\section{Experimental Evaluation}

\subsection{Experimental Design}

Three experiments were designed to validate EADD's detection accuracy, feature attribution, and computational efficiency. Following the evaluation protocol of Lukats et al.~[10], EADD was benchmarked against seven state-of-the-art unsupervised detectors: D3~[9], BNDM, CSDDM~[15], IBDD~[16], OCDD~[17], SPLL~[18], and UDetect.

\textbf{Datasets.} Table~\ref{tab:datasets} summarises the datasets used across experiments.

\begin{table}[htbp]
    \caption{Datasets Used in Experiments}
    \label{tab:datasets}
    \centering
    \begin{tabular}{|l|c|c|c|}
        \hline
        \textbf{Dataset} & \textbf{$n$} & \textbf{$d$} & \textbf{Drift?} \\
        \hline
        \multicolumn{4}{|l|}{\textit{Synthetic (Experiments 1, 3)}} \\
        \hline
        Temporal Patterns & 10K & 5 & Yes (4 types) \\
        Feature Attribution & 10K & 10 & Yes (3 scenarios) \\
        \hline
        \multicolumn{4}{|l|}{\textit{Real-World (Experiment 2)~[10],~[25]}} \\
        \hline
        INSECTS Abrupt Bal. & 52.8K & 33 & Yes \\
        INSECTS Gradual Bal. & 24.2K & 33 & Yes \\
        SineClusters & 10K & 2 & Yes \\
        WaveformDrift2 & 10K & 21 & Yes \\
        Electricity & 45.3K & 8 & Unknown \\
        NOAA Weather & 18.2K & 8 & Unknown \\
        Powersupply & 29.9K & 2 & Unknown \\
        \hline
    \end{tabular}
\end{table}

\textbf{Experiment 1 (Synthetic Data):} Synthetic streams ($n{=}10{,}000$, $d{=}5$) with controlled abrupt, gradual, incremental, and recurring drift injected at $t{=}5{,}000$. All eight detectors were compared. Five independent runs per drift type. Additionally, false alarm robustness was evaluated on four stable synthetic streams (Gaussian i.i.d., autocorrelated AR(1) with $\phi{=}0.8$, heteroscedastic, correlated with $\rho{=}0.7$) containing zero drift.

\textbf{Experiment 2 (Real-World Benchmark):} 11 real-world datasets from the USP Data Stream Repository~[25] following the Lukats et al.~[10] evaluation protocol: datasets with known drift points are evaluated by Mean Time to Detection (MTD) and Missed Drift Rate (MDR); datasets without ground truth are evaluated by total number of detections (selectivity).

\textbf{Experiment 3 (Feature Attribution Accuracy):} Three controlled scenarios using synthetic data with $d{=}10$ features and drift injected at $t{=}5{,}000$: (a)~univariate drift in F3 only (mean shift $+2\sigma$), (b)~subset drift in F2/F5/F7, (c)~multivariate drift in all features. Precision@k (P@k) and Recall@k (R@k) measure whether the top-$k$ SHAP-ranked features exactly match the $k$ truly drifted features.

\textbf{Configuration.} EADD: LightGBM~[23] (30 estimators, max depth 6, learning rate 0.1), $N_{ref}{=}500$, $N_{cur}{=}200$, monitoring frequency $M{=}50$, ASPT ($B_{\max}{=}50$, $B_{\min}{=}10$, $\alpha{=}0.01$, AUC threshold $\tau{=}0.7$). D3~[9]: logistic regression (liblinear solver), matching window sizes, fixed AUC threshold 0.7. All experiments: Python 3.10, LightGBM 4.1.0, SHAP~[24] 0.42.1, scikit-learn 1.3.2. Random seeds fixed for reproducibility.

% ===== RESULTS =====
\subsection{Experiment 1: Synthetic Data}

Table~\ref{tab:exp1} presents drift type coverage and false alarm rates for all eight detectors.

\begin{table}[htbp]
    \caption{Experiment 1: Synthetic Drift Detection Results (5 runs per type). Types = number of drift patterns detected out of 4. Mean FA = average false alarms per type. Lift = detections per true drift point (higher is better), following Lukats et al.~[10] reporting.}
    \label{tab:exp1}
    \centering
    \begin{tabular}{|l|c|c|c|c|}
        \hline
        \textbf{Detector} & \textbf{Types} & \textbf{Coverage} & \textbf{Mean FA} & \textbf{Lift} \\
        \hline
        \textbf{EADD} & 4/4 & 100\% & \textbf{0.25} & \textbf{1.06} \\
        BNDM & 4/4 & 100\% & 4.75 & 5.75 \\
        CSDDM & 4/4 & 100\% & 5.50 & 6.50 \\
        IBDD & 4/4 & 100\% & 46.25 & 47.25 \\
        OCDD & 4/4 & 100\% & 17.50 & 18.50 \\
        D3 & 2/4 & 50\% & 0.00 & 1.00 \\
        SPLL & 2/4 & 50\% & 0.00 & 1.00 \\
        UDetect & 0/4 & 0\% & --- & --- \\
        \hline
    \end{tabular}
\end{table}

EADD is the \textit{only} detector that achieved complete drift type coverage (4/4) with near-zero false alarms (mean 0.25 per type), yielding the best lift-per-drift ratio (1.06) among full-coverage detectors---i.e., nearly one detection per true drift point with minimal excess. Five detectors achieved 4/4 coverage, but with substantially higher false alarm costs: IBDD averaged 46.25, OCDD 17.5, CSDDM 5.5, and BNDM 4.75 false alarms per type. D3 and SPLL detected only 2/4 types, failing on gradual and incremental drift. UDetect failed on all four types.

\textbf{False Alarm Robustness.} On four stable streams containing zero drift, EADD produced near-zero false alarms (3.2 total across all stream types; zero on Gaussian, autocorrelated, and correlated streams). D3 was particularly vulnerable to autocorrelated data (58.8 false alarms at $\tau{=}0.7$). The ASPT correctly identifies temporal autocorrelation as non-significant ($p > 0.01$). A Mann--Whitney U test confirmed that EADD produces significantly fewer false alarms than D3 ($p = 0.013$).

\subsection{Experiment 2: Real-World Benchmark}

Table~\ref{tab:exp2} presents results on datasets with known drift points. EADD achieved the best MTD on WaveformDrift2 (99 vs D3's 179, a 44.7\% improvement) with zero false alarms. On 7 no-ground-truth datasets, EADD produced the fewest total detections (67) among functional detectors, indicating superior selectivity. IBDD produced 1{,}672 excess detections. BNDM timed out on all 33-feature datasets.

\begin{table}[htbp]
    \caption{Experiment 2: Real-World Benchmark (Datasets with Known Drift Points). MTD = Mean Time to Detection. MDR = Missed Drift Rate. $H_{\text{pred}}$ = mean prediction entropy at detection. ASPT = mean permutations used. ``---'' = missed all drifts or timed out.}
    \label{tab:exp2}
    \centering
    \begin{tabular}{|l|c|c|c|c|c|c|}
        \hline
        \textbf{Dataset} & \multicolumn{2}{c|}{\textbf{MTD}} & \multicolumn{2}{c|}{\textbf{MDR}} & $\boldsymbol{H_{\text{pred}}}$ & \textbf{ASPT} \\
        \cline{2-5}
         & \textbf{EADD} & \textbf{D3} & \textbf{EADD} & \textbf{D3} & & \textbf{Perms} \\
        \hline
        InsectsAbrupt & 173 & 153 & 0.0 & 0.0 & 0.21 & 11 \\
        InsectsGradual & --- & --- & 1.0 & 1.0 & 0.62 & --- \\
        SineClusters & 266 & 219 & 0.0 & 0.0 & 0.35 & 14 \\
        WaveformDrift2 & \textbf{99} & 179 & 0.0 & 0.0 & 0.18 & 10 \\
        \hline
    \end{tabular}
\end{table}

\subsection{Experiment 3: Feature Attribution Accuracy}

Table~\ref{tab:exp3} presents the SHAP attribution accuracy using precision@k and recall@k metrics.

\begin{table}[htbp]
    \caption{Experiment 3: SHAP Feature Attribution Accuracy. All scenarios detected at step 5{,}149 (149 samples after drift onset).}
    \label{tab:exp3}
    \centering
    \begin{tabular}{|l|c|c|c|c|c|}
        \hline
        \textbf{Scenario} & \textbf{AUC} & \textbf{P@k} & \textbf{R@k} & \textbf{Top (\%)} & \textbf{Rx} \\
        \hline
        Univariate & 0.784 & 1.0 & 1.0 & F3 (52.3) & $\checkmark$ \\
        Subset & 0.809 & 1.0 & 1.0 & F5 (24.5) & $\checkmark$ \\
        Multivariate & 0.806 & 1.0 & 1.0 & F2 (14.6) & $\checkmark$ \\
        \hline
    \end{tabular}
\end{table}

EADD achieved \textit{perfect} attribution accuracy: P@k~=~R@k~=~F1@k~=~1.0 across all three scenarios---the top-$k$ SHAP-ranked features exactly matched the truly drifted features in every case. The DSI-based prescription was correct in 3/3 cases (100\%). Among all eight detectors evaluated in Experiments~1--2, EADD is the only detector that identifies which features caused drift. In the univariate scenario, every other detector would require manual inspection of all 10~features, while EADD immediately pinpoints F3 at 52.3\% SHAP importance.

\subsection{Ablation Study}

To validate the contribution of each component, we compare EADD against D3 and systematically remove individual EADD components on the Experiment~1 synthetic streams (Table~\ref{tab:ablation}).

\begin{table}[htbp]
    \caption{Ablation Study: Building from D3 Baseline. Each row removes one EADD component. Perms = mean permutations per test (``---'' = no permutation test).}
    \label{tab:ablation}
    \centering
    \begin{tabular}{|l|c|c|c|}
        \hline
        \textbf{Configuration} & \textbf{Types} & \textbf{Mean FA} & \textbf{Perms} \\
        \hline
        D3~[9] (LogReg, fixed $\tau$) & 2/4 & 0.00 & --- \\
        \hline
        \multicolumn{4}{|l|}{\textit{Effect of removing EADD components:}} \\
        \hline
        $-$LightGBM ($\to$LogReg) & 2/4 & 0.00 & 12.8 \\
        $-$Reservoir ($\to$sliding) & 3/4 & 1.50 & 13.1 \\
        $-$ASPT ($\to$fixed $B_{\max}$) & 4/4 & 0.25 & 50.0 \\
        \hline
        \textbf{Full EADD} & \textbf{4/4} & \textbf{0.25} & \textbf{12.3} \\
        \hline
    \end{tabular}
\end{table}

Starting from D3's architecture, each EADD component addresses a specific limitation. Replacing D3's logistic regression with LightGBM is the most impactful change, improving coverage from 2/4 to 4/4 by capturing non-linear feature interactions in gradual and incremental drift. Reservoir sampling prevents reference window contamination by drifted data, recovering coverage from 3/4 to 4/4 and reducing false alarms. ASPT provides a $4\times$ computational reduction (50~$\to$~12.3 mean permutations) without affecting detection accuracy. Prediction entropy ($H_{\text{pred}}$) provides early-warning value for emerging drift that is not captured by binary detection metrics.

% ===== CONCLUSION =====
\section{Conclusion}

This paper presented EADD, a framework that advances discriminative drift detection through four methodological contributions: prediction entropy monitoring for early warning, Adaptive Sequential Permutation Testing for computationally efficient significance calibration, SHAP-based feature drift localisation for identifying which features cause drift, and the Drift Severity Index for continuous severity quantification with automated prescriptions.

Evaluation against seven state-of-the-art detectors demonstrated that EADD: (1)~is the only detector achieving complete drift coverage (4/4 types) with near-zero false alarms (mean 0.25), (2)~achieves 44.7\% faster detection on WaveformDrift2 with zero false alarms, (3)~achieves perfect feature attribution accuracy (P@k\,{=}\,R@k\,{=}\,1.0) with 100\% prescription accuracy, and (4)~produces significantly fewer false alarms than D3 ($p = 0.013$). By answering ``Is drift emerging?'', ``Is it significant?'', ``Which features caused it?'', and ``How severe is it?'', EADD transforms binary drift alarms into actionable diagnostic reports.

\subsection{Limitations and Future Work}

ASPT reduces computational cost by 70\% compared to fixed-budget permutation testing, but the adversarial paradigm remains ${\sim}50\times$ costlier than threshold-based D3. The DSI severity tiers ($0.3$, $0.6$) require domain-specific calibration. SHAP attributions are correlational, not causal~[33]. Future work includes e-value sequential testing~[27], causal drift attribution~[33], and adversarial robustness evaluation.

% ===== ACKNOWLEDGEMENTS (Optional) =====
% IMPORTANT: Comment out for blind review. Uncomment ONLY for camera-ready.
% -----------------------------------------------------------------------
% \section*{Acknowledgements}
%
% This research was conducted as part of a Master's thesis at the Faculty of
% Information and Communication Technology, Mahidol University, Thailand.
% -----------------------------------------------------------------------

% ===== REFERENCES =====
\section*{References}
\begin{small}
\begin{enumerate}[label={[\arabic*]}, leftmargin=*, itemsep=1pt, parsep=0pt]
    \item J.~Lu, A.~Liu, F.~Dong, F.~Gu, J.~Gama, and G.~Zhang, ``Learning under concept drift: A review,'' \textit{IEEE Trans. Knowl. Data Eng.}, vol.~31, no.~12, pp.~2346--2363, 2018.

    \item J.~Gama, I.~\v{Z}liobait\.{e}, A.~Bifet, M.~Pechenizkiy, and A.~Bouchachia, ``A survey on concept drift adaptation,'' \textit{ACM Comput. Surv.}, vol.~46, no.~4, pp.~1--37, 2014.

    \item G.~I. Webb, R.~Hyde, H.~Cao, H.~L. Nguyen, and F.~Petitjean, ``Characterizing concept drift,'' \textit{Data Min. Knowl. Discov.}, vol.~30, no.~4, pp.~964--994, 2016.

    \item F.~Bayram, B.~S.~Ahmed, and A.~Kassler, ``From concept drift to model degradation: An overview on performance-aware drift detectors,'' \textit{Knowl.-Based Syst.}, vol.~245, 108632, 2022.

    \item G.~Widmer and M.~Kubat, ``Learning in the presence of concept drift and hidden contexts,'' \textit{Mach. Learn.}, vol.~23, no.~1, pp.~69--101, 1996.

    \item S.~Rabanser, S.~G\"{u}nnemann, and Z.~Lipton, ``Failing loudly: An empirical study of methods for detecting dataset shift,'' in \textit{Proc. NeurIPS}, 2019, pp.~1396--1408.

    \item C.~Raab, M.~Heusinger, and F.-M.~Schleif, ``Reactive soft prototype computing for concept drift streams,'' \textit{Neurocomputing}, vol.~416, pp.~340--351, 2020.

    \item J.~Gama, P.~Medas, G.~Castillo, and P.~Rodrigues, ``Learning with drift detection,'' in \textit{Proc. Brazilian Symp. Artif. Intell.}, 2004, pp.~286--295.

    \item \"{O}.~G\"{o}z\"{u}a\c{c}{\i}k, A.~B\"{u}y\"{u}k\c{c}ak{\i}r, H.~Bonab, and F.~Can, ``Unsupervised concept drift detection with a discriminative classifier,'' in \textit{Proc. ACM CIKM}, 2019, pp.~2365--2368.

    \item D.~Lukats, O.~Zielinski, A.~Hahn, and F.~Stahl, ``A benchmark and survey of fully unsupervised concept drift detectors on real-world data streams,'' \textit{Int. J. Data Sci. Analytics}, vol.~19, pp.~1--31, 2024.

    \item M.~Baena-Garc\'{i}a et al., ``Early drift detection method,'' in \textit{Proc. ECML PKDD Workshop on Knowledge Discovery from Data Streams}, 2006, pp.~77--86.

    \item A.~Bifet and R.~Gavald\`{a}, ``Learning from time-changing data with adaptive windowing,'' in \textit{Proc. SIAM Int. Conf. Data Mining}, 2007, pp.~443--448.

    \item J.~Pan, V.~Pham, M.~Dorairaj, H.~Chen, and J.-Y.~Lee, ``Adversarial validation approach to concept drift problem in user targeting automation systems at Uber,'' in \textit{Proc. AdKDD Workshop, KDD}, 2020.

    \item H.~Qian, B.~Wang, P.~Ma, L.~Peng, S.~Gao, and Y.~Song, ``Managing dataset shift by adversarial validation for credit scoring,'' \textit{arXiv preprint arXiv:2112.10078}, 2021.

    \item J.~S.~Wan and S.-D.~Wang, ``Concept drift detection based on pre-clustering and statistical testing,'' \textit{J. Internet Technol.}, vol.~22, no.~2, pp.~275--286, 2021.

    \item V.~M.~A.~Souza, A.~R.~S.~Parmezan, F.~A.~Chowdhury, and A.~Mueen, ``Efficient unsupervised drift detector for fast and high-dimensional data streams,'' \textit{Knowl. Inf. Syst.}, vol.~63, no.~6, pp.~1497--1527, 2021.

    \item \"{O}.~G\"{o}z\"{u}a\c{c}{\i}k and F.~Can, ``Concept learning using one-class classifiers for implicit drift detection in evolving data streams,'' \textit{Artif. Intell. Rev.}, vol.~54, no.~5, pp.~3725--3747, 2021.

    \item L.~I.~Kuncheva, ``Change detection in streaming multivariate data using likelihood detectors,'' \textit{IEEE Trans. Knowl. Data Eng.}, vol.~25, no.~5, pp.~1175--1180, 2013.

    \item D.~Lopez-Paz and M.~Oquab, ``Revisiting classifier two-sample tests,'' in \textit{Proc. ICLR}, 2017.

    \item A.~Taly, K.~Sato, and C.~Gruia, ``Monitoring feature attributions: How Google saved one of the largest ML services in trouble,'' \textit{Google Cloud Blog}, 2021.

    \item AWS, ``Amazon SageMaker Clarify: Model monitoring,'' 2025. [Online]. Available: https://docs.aws.amazon.com/sagemaker

    \item J.~S. Vitter, ``Random sampling with a reservoir,'' \textit{ACM Trans. Math. Softw.}, vol.~11, no.~1, pp.~37--57, 1985.

    \item G.~Ke et al., ``LightGBM: A highly efficient gradient boosting decision tree,'' in \textit{Proc. NeurIPS}, 2017, pp.~3146--3154.

    \item S.~M. Lundberg and S.-I.~Lee, ``A unified approach to interpreting model predictions,'' in \textit{Proc. NeurIPS}, 2017, pp.~4765--4774.

    \item V.~M. A. Souza et al., ``Challenges in benchmarking stream learning algorithms with real-world data,'' \textit{Data Min. Knowl. Discov.}, vol.~34, pp.~1805--1858, 2020.

    \item T.~S.~Sethi and M.~Kantardzic, ``On the reliable detection of concept drift from streaming unlabeled data,'' \textit{Expert Syst. Appl.}, vol.~82, pp.~77--99, 2017.

    \item T.~Pandeva, T.~Bakker, C.~A.~Naesseth, and P.~Forr\'{e}, ``E-valuating classifier two-sample tests,'' \textit{Trans. Mach. Learn. Res.}, 2024.

    \item P.~Panda, S.~S.~Kancheti, V.~N.~Balasubramanian, and G.~Sinha, ``Interpretable model drift detection,'' in \textit{Proc. ACM IKDD CODS-COMAD}, 2024.

    \item J.~Feng et al., ``HDPD: Hierarchical decomposition of performance discrepancy,'' in \textit{Proc. NeurIPS}, 2024.

    \item A.~Gandy, ``Sequential implementation of Monte Carlo tests with uniformly bounded resampling risk,'' \textit{J. Am. Stat. Assoc.}, vol.~104, no.~488, pp.~1504--1511, 2009.

    \item S.~M.~Lundberg, G.~Erion, H.~Chen et al., ``From local explanations to global understanding with explainable AI for trees,'' \textit{Nature Mach. Intell.}, vol.~2, pp.~56--67, 2020.

    \item F.~Hinder, V.~Vaquet, and B.~Hammer, ``Feature-based analyses of concept drift,'' \textit{Neurocomputing}, vol.~600, 127968, 2024.

    \item K.~Budhathoki, D.~Janzing, P.~Bloebaum, and H.~Ng, ``Why did the distribution change?'' in \textit{Proc. AISTATS}, 2021, pp.~1666--1674.

    \item C.~E. Shannon, ``A mathematical theory of communication,'' \textit{Bell Syst. Tech. J.}, vol.~27, no.~3, pp.~379--423, 1948.
\end{enumerate}
\end{small}

% ===== ABOUT THE AUTHORS =====
% IMPORTANT: Comment out this entire section for blind review submission.
% Uncomment ONLY for final camera-ready version after acceptance.
% -----------------------------------------------------------------------
% \section*{About the Authors}
%
% \textbf{Nusrat Begum} is a Master's student in Computer Science at
% the Faculty of ICT, Mahidol University, Thailand. Research interests
% include MLOps, concept drift detection, and explainable AI.
%
% \vspace{6pt}
%
% \textbf{Thanapon Noraset} is an Assistant Professor at the Faculty
% of ICT, Mahidol University. Research focus includes natural language
% processing and machine learning.
%
% \vspace{6pt}
%
% \textbf{Suppawong Tuarob} is an Associate Professor at the Faculty
% of ICT, Mahidol University. Specializes in data mining and information
% retrieval.
% -----------------------------------------------------------------------

% Balance columns on last page
\balance

\end{document}
